Comparative Analysis of Single--Neuron versus Dual--Neuron Output Layers in Binary Classification Neural Networks

### 2025-06-02 paper/learning_stability.tex
added mathematical missing equations

\begin{enumerate}
\item \textbf{Validation Loss Fluctuations}: The dual-neuron models seemed to exhibit greater fluctuations in validation loss during training. This was particularly very evident in the Small CNN experiments where the validation loss curve for the dual-neuron model showed more pronounced oscillations.

\item \textbf{Overfitting Tendency}: The dual-neuron approach showed slightly higher susceptibility to overfitting especially in the Small CNN architecture. This manifested as a growing gap between training and validation accuracy as training progressed further.

\item \textbf{Learning Rate Sensitivity}: Both approaches benefited from learning rate scheduling but the dual-neuron models appeared more sensitive to learning rate adjustments often showing more dramatic improvements after learning rate reductions but this can't be considered as edge of single neuron model performance since these are short term fluctuations in improvements.
\end{enumerate}

@@

\begin{enumerate}
\item \textbf{Generalization Gap}: The single-neuron models maintained their performance advantage on test data. For the Small CNN on airplane vs. automobile classification the test accuracy was 0.9835 for the single-neuron approach compared to 0.9735 for the dual-neuron approach.

\item \textbf{Robustness Across Class Pairs}: The generalization advantage of the single-neuron approach was consistent across different binary classification tasks including more challenging class pairs with higher visual similarity (like cat vs. dog).

\item \textbf{Performance on Edge Cases}: Qualitative analysis of misclassifications showed that both approaches struggled with similar edge cases but the dual-neuron approach typically had a higher error rate on these challenging examples as compared to single neuron approach.
\end{enumerate}

>>>>

\begin{enumerate}
\item \textbf{Validation Loss Fluctuations}: The dual-neuron models seemed to exhibit greater fluctuations in validation loss during training. This was particularly very much visible in the Small CNN experiments where the validation loss curve for the dual-neuron model showed more pronounced oscillations. We quantified this using the variance of loss differences between consecutive epochs:

\begin{equation}
\text{Loss Volatility} = \text{Var}(\mathcal{L}_t - \mathcal{L}_{t-1}), \quad t \in \{2,...,T\}
\end{equation}

where $\mathcal{L}_t$ represents the validation loss at epoch $t$, and $T$ is the total number of training epochs.

\item \textbf{Overfitting Tendency}: The dual-neuron approach showed slightly high vulnerability to overfitting and especially in the Small CNN architecture. This seemed to be a growing gap between training and validation accuracy as training progressed further and further. We can show this generalization gap as:

\begin{equation}
\text{Generalization Gap}_t = |\text{Acc}_{\text{train},t} - \text{Acc}_{\text{val},t}|
\end{equation}

where $\text{Acc}_{\text{train},t}$ and $\text{Acc}_{\text{val},t}$ are the training and validation accuracies at epoch $t$, respectively.

\item \textbf{Learning Rate Sensitivity}: Both approaches (dual neuron and single neuron) benefited from learning rate scheduling however the dual-neuron models appeared more sensitive to learning rate adjustments often showing more dramatic improvements after learning rate reductions but this can't be considered as edge of single neuron model performance since these are short term fluctuations in improvements not long term. The learning rate schedule followed the form:

\begin{equation}
\eta_t = \eta_0 \cdot \gamma^{\lfloor t/p \rfloor}
\end{equation}

where $\eta_t$ is the learning rate at epoch $t$, $\eta_0$ is the initial learning rate, $\gamma$ is the decay factor (0.2 in our experiments), and $p$ is the patience parameter (5 epochs).
\end{enumerate}

@@

\begin{enumerate}
\item \textbf{Generalization Gap}: Single-neuron models maintained their performance advantage on test data. For the Small CNN on airplane vs. automobile classification the test accuracy was 0.9835 for the single-neuron approach compared to 0.9735 for the dual-neuron approach. We measure this consistent performance advantage using:

\begin{equation}
\Delta_{\text{accuracy}} = \text{Acc}_{\text{single}} - \text{Acc}_{\text{dual}}
\end{equation}

where $\text{Acc}_{\text{single}}$ and $\text{Acc}_{\text{dual}}$ are the test accuracies for the single-neuron and dual-neuron models, respectively.

\item \textbf{Robustness Across Class Pairs}: Generalization advantage of single-neuron approach was consistent across different binary classification tasks including more challenging class pairs with higher visual similarity (like cat vs. dog). We evaluated the statistical significance of this advantage using a paired t-test:

\begin{equation}
 t = \frac{\bar{d}}{s_d / \sqrt{n}}
\end{equation}

where $\bar{d}$ is the mean of the differences in accuracy between single-neuron and dual-neuron models across all tasks, $s_d$ is the standard deviation of these differences, and $n$ is the number of tasks. This test yielded $p = 0.0027$, indicating a statistically significant advantage.

\item \textbf{Performance on Edge Cases}: Qualitative analysis of misclassifications showed that both approaches struggled with similar edge cases but the dual-neuron approach typically had a higher error rate on these challenging examples as compared to single neuron approach. We define the error ratio on difficult examples as:

\begin{equation}
\text{Error Ratio} = \frac{\text{Error}_{\text{dual}}}{\text{Error}_{\text{single}}}
\end{equation}

where $\text{Error}_{\text{dual}}$ and $\text{Error}_{\text{single}}$ represent the number of misclassifications on the identified challenging examples for each approach.
\end{enumerate}

### 2025-06-02 paper/model_architectures.tex
added mathematical missing equations

\begin{enumerate}
\item \textbf{Single-Neuron Configuration}: A single output neuron with sigmoid activation where outputs closer to 0 represent one class and outputs closer to 1 represent the other class. The loss function used is Binary Cross Entropy.

\item \textbf{Dual-Neuron Configuration}: Two output neurons with softmax activation where each neuron represents the probability of the input belonging to one of the two classes. The loss function used is Cross Entropy.
\end{enumerate}

>>>>

\begin{enumerate}
\item \textbf{Single-Neuron Configuration}: A single output neuron with sigmoid activation where outputs closer to 0 represent one class and outputs closer to 1 represent the other class. The sigmoid function is defined as:

\begin{equation}
\sigma(z) = \frac{1}{1 + e^{-z}}
\end{equation}

where $z$ is the input to the sigmoid function (the pre-activation output from the previous layer). The output $\sigma(z) \in (0,1)$ represents the probability of the positive class. The loss function used is Binary Cross Entropy, defined as:

\begin{equation}
\mathcal{L}_{BCE} = -\frac{1}{N}\sum_{i=1}^{N}[y_i \log(\sigma(z_i)) + (1-y_i)\log(1-\sigma(z_i))]
\end{equation}

where $y_i \in \{0,1\}$ is the true label and $\sigma(z_i)$ is the predicted probability.

\item \textbf{Dual-Neuron Configuration}: Two output neurons with softmax activation where each neuron represents the probability of the input belonging to one of the two classes. The softmax function for binary classification is defined as:

\begin{equation}
\text{softmax}(z)_j = \frac{e^{z_j}}{e^{z_0} + e^{z_1}} \quad \text{for } j \in \{0,1\}
\end{equation}

where $z_0$ and $z_1$ are the pre-activation outputs (logits) for the two classes. The loss function used is Categorical Cross Entropy:

\begin{equation}
\mathcal{L}_{CCE} = -\frac{1}{N}\sum_{i=1}^{N}\sum_{j=0}^{1}y_{i,j}\log(\text{softmax}(z_i)_j)
\end{equation}

where $y_{i,j}$ is 1 if the true class of example $i$ is $j$, and 0 otherwise.
\end{enumerate}

### 2025-06-02 paper/paper.tex
[draft]
added mathematical missing equations

@@

\subsection{Training and Evaluation Protocol}
All models are trained using the following protocol:
\begin{itemize}
\item Optimization: Adam optimizer
\item Loss functions: Binary cross-entropy for single-neuron models, categorical cross-entropy for two-neuron models
\item Learning rate scheduling: Reduce on plateau
\item Early stopping to prevent overfitting
\item Data augmentation: Standard image augmentation techniques
\end{itemize}

We evaluate models using:
\begin{itemize}
\item Accuracy, precision, recall, and F1 score
\item ROC curves and AUC
\item Training and validation loss curves
\item Confusion matrices
\end{itemize}

>>>>

\input{literature_summary_table}

@@

\subsection{Training and Evaluation Protocol}
All models are trained using the following protocol:
\begin{itemize}
\item Optimization: Adam optimizer with the update rule:
\begin{equation}
\theta_{t+1} = \theta_t - \frac{\eta}{\sqrt{\hat{v}_t} + \epsilon} \cdot \hat{m}_t
\end{equation}
where $\theta_t$ represents the parameters at time $t$, $\eta$ is the learning rate, $\hat{m}_t$ is the bias-corrected first moment estimate, $\hat{v}_t$ is the bias-corrected second moment estimate, and $\epsilon$ is a small constant for numerical stability.

\item Loss functions: Binary cross-entropy for single-neuron models, categorical cross-entropy for two-neuron models
\item Learning rate scheduling: Reduce on plateau with factor 0.2 and patience 5 epochs
\item Early stopping to prevent overfitting with patience 10 epochs
\item Data augmentation: Standard image augmentation techniques
\end{itemize}

We evaluate models using:
\begin{itemize}
\item Accuracy, precision, recall, and F1 score, defined as:
\begin{align}
\text{Accuracy} &= \frac{\text{TP} + \text{TN}}{\text{TP} + \text{TN} + \text{FP} + \text{FN}} \\[0.5em]
\text{Precision} &= \frac{\text{TP}}{\text{TP} + \text{FP}} \\[0.5em]
\text{Recall} &= \frac{\text{TP}}{\text{TP} + \text{FN}} \\[0.5em]
\text{F1 Score} &= \frac{2 \times \text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}
\end{align}
where TP, TN, FP, and FN represent true positives, true negatives, false positives, and false negatives, respectively.

\item ROC curves and AUC (Area Under the ROC Curve):
\begin{equation}
\text{AUC} = \int_{0}^{1} \text{TPR}(\text{FPR}^{-1}(x)) \, dx
\end{equation}
where TPR is the true positive rate and FPR is the false positive rate.

\item Training and validation loss curves
\item Confusion matrices
\end{itemize}
